Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation

### 2025-04-18 manuscript/manuscript.qmd
Cleaned introduction

### Introduction

#### Historical Context of Chlorophyll Algorithms
Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable the global estimation of chlorophyll‑a ($Chl_a$) —a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

@@

#### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability. Consider data $D$ that include predictive features (e.g. Remote sensing reflectance - Rrs) and prediction targets (e.g. $Chl_a$ or phytoplankton absorption), and a model $M$ (e.g. OCx) hypothesized to adequately represent the statistical association between the two. The sampling probability $p(D|M)$ also referred to as the likelihood, is the probability of $D$ conditioned on $M$ being "true". The common model fitting practice is to maximize $p(D|M)$ - likelihood maximization - achieved by tuning $M$'s parameters. The unspoken assumption is that making $M$ fit $D$ as well as possible yields the best version of $M$ given $D$ i.e.  maximizing $p(M|D)$, the inferential probability. This sneaky swapping from $p(D|M)$ to $p(M|D)$ is a violation of the rules of conditional probability, Bayes' theorem in particular. Jaynes[@jaynes2003] and later Clayton through his highly readable account [@clayton] are some of the more notable whistleblowers. Occasionally, in simple  or well-behaved problems, with a large amount of data and no prior information available the maxima of both $p(M|D)$ and $p(D|M)$ can indeed converge on the same answer. Yet there is no guarantee predictability for this outcome however, and the fallout of this abuse is non-negligible, most noticeably in terms of models that don't generalize well, ad-hoc uncertainty estimation attempts, and more broadly a replication crisis.

#### Overcoming limitations
Some researchers have attempted to deal with these issues. [@seegers2018] e.g. have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher and switch to the Bayesian paradigm. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout as deep neural net training method to obtain prediction uncertainty. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a forward model. [@hammout2024] have proposed a Bayesian Neural Network trained using Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].

#### In search of the Ocean Color Bayesian workflow
The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g. [@mcelreath_stat_rethink]). That is not to say that all steps are applicable to all fields of research and all intents. The workflow can and should be tailored to the researcher's field. [@wolkovich2024fourstepbayesianworkflowimprovin] has proposed a four-step process for Ecological modeling. 
However there are some core steps that should be followed. (1) Begin with one, or better yet,  more than one conceptual models and code them up.  Model building include prior formulation to encode existing knowledge. Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvemnt. (2) Check assumptions by a simulation process known as *Prior Predictive Checks*. Bayesian models are generative, meaning they can run on empty, that is produce results without data. This allows a sanity check of the built-in assumptions. Non-sensical simulation output indicate assumptions must be revisited. (3) Collect data and conduct exploratory data analysis. Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features. (4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information. Most modern packages (e.g. Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights. (5) Model compasison and selection. Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights. (6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.

In this paper, we largely follow this workflow. We begin by leveraging some of the work eluded to above and  recast some of the pre-existing $OCx$ models into their probabilistic version. We also offer an alternative in the form of Bayesian Additive Regression Trees  (BART; [@chipman2010bart]).

### Methods

#### Prior Elicitation and Model Formulation

#### Data Preprocessing

####

>>>>

## Introduction

### Historical Context of Chlorophyll Algorithms
Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

@@

### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g. Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them. The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true". Standard model fitting maximizes this quantity (i.e. the likelihood) by tuning the parameters of $M$.

The unspoken assumption is that the model that best fits the data also best represents the underlying process. This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem. While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule. The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.

In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]). These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well. In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.

This epistemic shortcut has already been explicitly critiqued in the machine learning community. Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference. This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs. Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.

### Overcoming limitations
There have been attempts to address these issues. [@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher incorporate Bayesian concepts. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout to approximate a deep BNN. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty. [@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].

### In search of our Bayesian workflow
The Bayesian workflow is a recognizable and unifying way of approaching statistical modeling, initially championed by statisticians at Columnbia University [@gelman_bda] subsequently quick endorsed and further developed by researchers dealing with scant and noisy data (e.g. [@mcelreath_stat_rethink]). That is not to say that all steps are applicable to all fields of research and all intents. The workflow can and should be tailored to the researcher's field. [@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for Ecological modeling. 
However there are some core steps that should be followed. (1) Begin with one, or better yet,  more than one conceptual models and code them up.  Model building include prior formulation to encode existing knowledge. Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement. (2) Check assumptions by a simulation process known as *Prior Predictive Checks*. Bayesian models are generative, meaning they can run on empty, that is produce results without data. This allows a sanity check of the built-in assumptions. Non-sensical simulation results indicate assumptions must be revisited. (3) Collect data and conduct exploratory data analysis. Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features. (4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information. Most modern packages (e.g. Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights. (5) Model compasison and selection. Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights. (6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.

In this paper we leverage past work and recast some of the pre-existing $OCx$ models into their probabilistic version. We also propose alternative models, namely Bayesian Additive Regression Trees (BART, [@chipman2010bart]) as a robust flexible method that has been used for diverse application ([@linero2017].)

## Methods

### Prior Elicitation and Model Formulation

### Data Preprocessing

###

### 2025-04-18 manuscript/manuscript.tex
Cleaned introduction

\authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
\affiliation{1}{Consultant, }\affiliation{2}{NASA, }
\correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}

@@

\subsection{Introduction}\label{introduction}

\subsubsection{Historical Context of Chlorophyll
Algorithms}\label{historical-context-of-chlorophyll-algorithms}

@@

\subsubsection{Limitations of Existing
Approaches}\label{limitations-of-existing-approaches}

Regrettably, the development of traditional ocean color predictive
models relies on a fundamental statistical error that plagues most of
observational science today; that of conflating sampling probability,
with inferential probability. Consider data \(D\) that include
predictive features (e.g.~Remote sensing reflectance - Rrs) and
prediction targets (e.g.~\(Chl_a\) or phytoplankton absorption), and a
model \(M\) (e.g.~OCx) hypothesized to adequately represent the
statistical association between the two. The sampling probability
\(p(D|M)\) also referred to as the likelihood, is the probability of
\(D\) conditioned on \(M\) being ``true''. The common model fitting
practice is to maximize \(p(D|M)\) - likelihood maximization - achieved
by tuning \(M\)'s parameters. The unspoken assumption is that making
\(M\) fit \(D\) as well as possible yields the best version of \(M\)
given \(D\) i.e.~maximizing \(p(M|D)\), the inferential probability.
This sneaky swapping from \(p(D|M)\) to \(p(M|D)\) is a violation of the
rules of conditional probability, Bayes' theorem in particular.
Jaynes(\textbf{jaynes2003?}) and later Clayton through his highly
readable account (\textbf{clayton?}) are some of the more notable
whistleblowers. Occasionally, in simple or well-behaved problems, with a
large amount of data and no prior information available the maxima of
both \(p(M|D)\) and \(p(D|M)\) can indeed converge on the same answer.
Yet there is no guarantee predictability for this outcome however, and
the fallout of this abuse is non-negligible, most noticeably in terms of
models that don't generalize well, ad-hoc uncertainty estimation
attempts, and more broadly a replication crisis.

Some researchers have attempted to deal with these issues. (Seegers et
al., 2018) e.g.~have proposed alternative metrics to circumvent the
inadequate assumptions of the frequentist approach. Others have tried to
go a step futher and switch to the Bayesian paradigm. E.g. (Frouin \&
Pelletier, 2013) have proposed a Bayesian inversion scheme for
atmospheric correction. (\textbf{shi2015?}) proposed a probabilistic
method to merge data from different sensors. (Craig \& Karaköylü, 2019)
have proposed a Bayesian neural network approach using Hamiltonian Monte
Carlo sampling to retrieve Inherent Optical Properties (IOP) from
Top-of-the-atmosphere (TOA) radiance. (\textbf{werther2022?}) used
Monte-Carlo dropout as deep neural net training method to obtain
prediction uncertainty. (Erickson et al., 2023) have proposed using
conjugate Gaussian prior and likelihood to predict IOPs using GIOP as a
forward model. (\textbf{hammout2024?}) have proposed a Bayesian Neural
Network trained using Stochastic Variational Inference to predict
\(Chl_a\) from ocean color observations. Yet most of these approaches
retain variable levels of frequentism by applying only part of what is
commonly referred to as the Bayesian workflow(\textbf{betancourt2019?}).

In this paper we propose a modified workflow to Bayesian modeling for
Ocean Color emote sensing. We propose that this workflow include; (1)
one or more model building, (2) prior formulation via expert knowledge
elicitation and/or data-free simulation, (3) fitting model to data, (4)
model validation, comparison and selection, (5) interpretation of
results via posterior analysis, (6) prediction on new observation, (7)
update the posterior when new labeled data becomes available. As
illustration we recast some of the pre-existing \(OCx\) models into
their probabilistic version. To illustrate model comparison and
selection we also build and evaluate Bayesian Additive Tree Regregssion
(BART; (Chipman et al., 2010)) type model.

\subsection{Methods}\label{methods}

\subsubsection{Prior Elicitation and Model
Formulation}\label{prior-elicitation-and-model-formulation}

\subsubsection{Data Preprocessing}\label{data-preprocessing}

\subsubsection{}\label{section}

@@

@@

@@

@@

@@

@@

>>>>

\authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
\affiliation{1}{Independant Consultant, }\affiliation{2}{NASA, }
\correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}

@@

\section{Introduction}\label{introduction}

\subsection{Historical Context of Chlorophyll
Algorithms}\label{historical-context-of-chlorophyll-algorithms}

@@

\subsection{Limitations of Existing
Approaches}\label{limitations-of-existing-approaches}

Regrettably, the development of traditional ocean color predictive
models relies on a fundamental statistical error that plagues most of
observational science today; that of conflating sampling probability,
with inferential probability(Scheemaekere \& Szafarz (2011).) Consider a
dataset \(D\) that consists of input-output pairs - e.g.~Remote sensing
reflectance (Rrs) and chlorophyll-\emph{a} concentration \(Chl_a\) - and
a model \(M\), such as OCx, hypothesized to describe the statistical
association between them. The sampling probability \(p(D|M)\) represents
the probability of observing the data \(D\) if model \(M\) were
``true''. Standard model fitting maximizes this quantity (i.e.~the
likelihood) by tuning the parameters of \(M\).

The unspoken assumption is that the model that best fits the data also
best represents the underlying process. This constitutes an epistemic
fallacy; substituting \(p(D|M)\) for \(p(M|D)\), in a violation of the
rules of conditional probability inversion as framed by Bayes' theorem.
While in well-behaved data-rich problems, the maxima of both expressions
may coincide, this is an exception, not the rule. The consequences of
this mistake are far-reaching as argued by (Clayton, 2022), with
implications extending from scientific modeling to medicine, law, and
public policy.

In science, this fallacy contributes to models fail to generalize well,
an impetus to produce ad-hoc or post-hoc uncertainty quantification, and
yield published results that resist replication(Cobey et al. (2024)).
These limiations are not confined to traditional hypothesis testing;
they are embedded in the training.; they are embedded in the training
and deployment of modern machine learning models as well. In regression
and classification tasks, likelihood maximization is treated as
sufficient for inference, even though it results in a single point
estimate and fails to account for parameter uncertainty or model
plausibility.

This epistemic shortcut has already been explicitly critiqued in the
machine learning community. Gal (Gal, 2016) and Ghahramani
(\textbf{gahramani2015probabilistic?}) emphasize that most ML models
discard uncertainty entirely and treat the output of optimization as
inference. This leads to overconfident predctions and brittle
generalization, echoing Clayton's concerncs. Bishop (Bishop, 2006) also
distiguishes between the utility of models and the inferential core
required to quantify uncertainty, underscoring that likelihood is not
enough and that the Bernoulli fallacy pervades the very logic of applied
machine learning.

\subsection{Overcoming limitations}\label{overcoming-limitations}

There have been attempts to address these issues. (Seegers et al., 2018)
have proposed alternative metrics to circumvent the inadequate
assumptions of the frequentist approach. Others have tried to go a step
futher incorporate Bayesian concepts. E.g. (Frouin \& Pelletier, 2013)
have proposed a Bayesian inversion scheme for atmospheric correction.
(\textbf{shi2015?}) proposed a probabilistic method to merge data from
different sensors. (Craig \& Karaköylü, 2019) have proposed a Bayesian
neural network (BNN) approach using Hamiltonian Monte Carlo sampling to
retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere
(TOA) radiance. (\textbf{werther2022?}) used Monte-Carlo dropout to
approximate a deep BNN. (Erickson et al., 2023) have proposed using
conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian
model to predict IOPs with uncertainty. (\textbf{hammout2024?}) have
proposed a BNN approximation via Stochastic Variational Inference to
predict \(Chl_a\) from ocean color observations. Yet most of these
approaches retain variable levels of frequentism by applying only part
of what is now commonly referred to as the Bayesian
workflow(\textbf{bgelman2019?}).

\subsection{In search of our Bayesian
workflow}\label{in-search-of-our-bayesian-workflow}

The Bayesian workflow is a recognizable and unifying way of approaching
statistical modeling, initially championed by statisticians at Columnbia
University (\textbf{gelman\_bda?}) subsequently quick endorsed and
further developed by researchers dealing with scant and noisy data (e.g.
(\textbf{mcelreath\_stat\_rethink?})). That is not to say that all steps
are applicable to all fields of research and all intents. The workflow
can and should be tailored to the researcher's field. (Wolkovich et al.,
2024) has proposed a four-step process for Ecological modeling. However
there are some core steps that should be followed. (1) Begin with one,
or better yet, more than one conceptual models and code them up. Model
building include prior formulation to encode existing knowledge. Priors
are part of the model structure and make the researcher's assumptions
transparent, providing a first avenue of critique and improvement. (2)
Check assumptions by a simulation process known as \emph{Prior
Predictive Checks}. Bayesian models are generative, meaning they can run
on empty, that is produce results without data. This allows a sanity
check of the built-in assumptions. Non-sensical simulation results
indicate assumptions must be revisited. (3) Collect data and conduct
exploratory data analysis. Knowing the data will be critical to
understanding potential problems during fitting; e..g multicollinearity
of input features. (4) Diagnostics; the fitting process and the
resulting posterior distribution provide rich constructs that can be
mined for a great deal of information. Most modern packages (e.g.~Stan,
PyMC, Numpyro, etc) offer a great many tools to extract insights. (5)
Model compasison and selection. Often a model will perform markedly
better, which is useful in terms predictive performance, but more
importantly in terms of generating insights. (6) Sequential update; a
model's output, the posterior distribution becomes the new prior every
time new data becomes available for a new sequence of model fitting.

In this paper we leverage past work and recast some of the pre-existing
\(OCx\) models into their probabilistic version. We also propose
alternative models, namely Bayesian Additive Regression Trees (BART,
(Chipman et al., 2010)) as a robust flexible method that has been used
for diverse application ((Linero, 2017).)

\section{Methods}\label{methods}

\subsection{Prior Elicitation and Model
Formulation}\label{prior-elicitation-and-model-formulation}

\subsection{Data Preprocessing}\label{data-preprocessing}

\subsection{}\label{section}

@@

\bibitem[\citeproctext]{ref-baker2016}
Baker, M. (2016). 1,500 scientists lift the lid on reproducibility.
\emph{Nature}, \emph{533}(7604), 452--454.
\url{https://doi.org/10.1038/533452a}

\bibitem[\citeproctext]{ref-bishop2006pattern}
Bishop, C. M. (2006). \emph{Pattern recognition and machine learning}.
Springer.

@@

\bibitem[\citeproctext]{ref-clayton2022bernoulli}
Clayton, A. (2022). \emph{Bernoulli's fallacy: Statistical illogic and
the crisis of modern science}. Columbia University Press. Retrieved from
\url{https://books.google.com/books?id=BT4CzwEACAAJ}

\bibitem[\citeproctext]{ref-cobey2024biomedical}
Cobey, K. D., Ebrahimzadeh, S., Page, M. J., Thibault, R. T., Nguyen,
P.-Y., Abu-Dalfa, F., \& Moher, D. (2024). Biomedical researchers'
perspectives on the reproducibility of research. \emph{PLoS Biology},
\emph{22}(11), e3002870.

@@

\bibitem[\citeproctext]{ref-gal2016uncertainty}
Gal, Y. (2016). \emph{Uncertainty in deep learning} (PhD thesis).
University of Cambridge.

@@

\bibitem[\citeproctext]{ref-jaynes2003probability}
Jaynes, E. T., \& Bretthorst, G. L. (2003). \emph{Probability theory:
The logic of science}. Cambridge University Press. Retrieved from
\url{https://books.google.com/books?id=tTN4HuUNXjgC}

\bibitem[\citeproctext]{ref-linero2017}
Linero, A. R. (2017). A review of tree-based bayesian methods.
\emph{Communications for Statistical Applications and Methods},
\emph{24}, 543--559. \url{https://doi.org/10.29220/CSAM.2017.24.6.543}

@@

\bibitem[\citeproctext]{ref-DeScheemaekere2011}
Scheemaekere, X. D., \& Szafarz, A. (2011). The inference fallacy from
bernoulli to kolmogorov. Retrieved from
\url{https://www.researchgate.net/publication/254450993_The_Inference_Fallacy_From_Bernoulli_to_Kolmogorov}

@@

\bibitem[\citeproctext]{ref-wolkovich2024fourstepbayesianworkflowimproving}
Wolkovich, E., Davies, T. J., Pearse, W. D., \& Betancourt, M. (2024). A
four-step bayesian workflow for improving ecological science. Retrieved
from \url{https://arxiv.org/abs/2408.02603}
